The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-12-06 paper/main.tex
One more proofreading pass.

\begin{abstract}
Team cohesion and the dynamics of team formation are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. Because of these relationships, representing software team members as a graph may be a natural way to explore these relationships. As team members move between projects, these graphs become more and more connected as team members collaborate and form new relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\subsection{Hypothesis}
The formalization of the ``small world'' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, the probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance. Thus, we formulate our hypothesis:\\

@@

\subsection{Collection and Storage}
To explore our hypothesis, software project data was collected and analyzed. This data was collected using the freely available FLOSSmole\cite{floss2006} datasets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode},\footnote{http://freecode.com/} and another from open source software repository \textit{SourceForge}.\footnote{http://sourceforge.net/} The Freecode dataset contained data up to September 2013, while the SourceForge dataset was slightly older with data gathered up to June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing package NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

@@

\noindent\\\textit{\textbf{Average Path Length ($L$)}}\\
Calculating the \textit{Path Length} ($L$) allowed us to determine how far removed contributors in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. Exploring this phenomenon further, we can derive that most graphs resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.

@@

\noindent\\\textit{\textbf{New contributors are good, as long as the core team remains intact.}}\\
Both participants mentioned the idea of keeping a ``core" team intact. This was certainly an often cited notion in both interviews. This concept appears to align with the idea of a \textit{core/periphery}\cite{borgatti2000models}, where the graph under study contains one (or many) tightly interconnected subgraphs, with less tightly connected nodes radiating out from this core.

@@

\section{Conclusion}
Our hypothesis theorizes that medium values of clustering between collaborators on projects is optimal for project performance. The data analysis, specifically figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}, seem to support this claim. Also, qualitative research seems to support the hypothesis as well. Our interviews revealed that from experience, teams perform best when a stable core is periodically augmented with new members. This would be a qualitative representation of a intermediate $Q$ value -- significant clustering but not to the extent of isolation. Given both the quantitative and qualitative analysis, we can assert to align with Uzzi's findings\cite{uzzi2005collaboration} and show that intermediate values of $Q$ are correlated with increased project performance.

>>>>

\begin{abstract}
Team cohesion and the dynamics of team formation are important parts of any project, with software projects being no exception. An interesting aspect of team building is the relationships formed between the team members. Because of these relationships, representing software team members as a graph may be a natural way to explore team dynamics. As team members move between projects, these graphs become more and more connected as team members collaborate and form new relationships. We show that this connectivity, known as the ``small world effect," has a positive impact on team performance when the connectivity levels are moderate. Performance degrades, however, at both very high and very low levels of connectivity. This aligns with similar research findings of non-software teams.
\end{abstract}

@@

\subsection{Hypothesis}
The formalization of the ``small world'' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of a graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, the probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance. Thus, we formulate our hypothesis:\\

@@

\subsection{Collection and Storage}
To explore our hypothesis, software project data was collected and analyzed. This data was collected using the freely available FLOSSmole\cite{floss2006} datasets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode},\footnote{http://freecode.com/} and another from the open source software repository \textit{SourceForge}.\footnote{http://sourceforge.net/} The Freecode dataset contained data up to September 2013, while the SourceForge dataset was slightly older with data gathered up to June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing package NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

@@

\noindent\\\textit{\textbf{Average Path Length ($L$)}}\\
Calculating the \textit{Average Path Length} ($L$) allowed us to determine how far removed contributors in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power-law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. Exploring this phenomenon further, we can derive that most graphs conceptually resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.

@@

\noindent\\\textit{\textbf{New contributors are good, as long as the core team remains intact.}}\\
Both participants mentioned the idea of keeping a ``core" team intact, and was an often cited concept in both interviews. This concept appears to align with the idea of a \textit{core/periphery}\cite{borgatti2000models}, where the graph under study contains one (or many) tightly interconnected subgraphs, with less tightly connected nodes radiating out from this core.

@@

\section{Conclusion}
Our hypothesis theorizes that medium values of clustering between collaborators on projects is optimal for project performance. The data analysis, specifically figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}, seem to support this claim. Also, our qualitative research tends to support the hypothesis as well. Our interviews revealed that from experience, teams perform best when a stable core is periodically augmented with new members. This would be a qualitative representation of a intermediate $Q$ value -- significant clustering but not to the extent of isolation. Given both the quantitative and qualitative analysis, we can assert to align with Uzzi's findings\cite{uzzi2005collaboration} and show that intermediate values of $Q$ are correlated with increased project performance.
